Manuscript

### 2025-09-21 Manuscript_MD_ScoreReLiability.qmd
update manuscript up to appendix

@@

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($\delta_0$) and corrected ES
($\delta$), which are central to assess the research questions posed in
section *Attenuation Correction*. Most importantly, @tbl-tauES
demonstrates that for all but one of the phenomena, the attenuation
correction procedure barely has an impact on the extent of ES
heterogeneity. However, while the heterogeneity for the phenomenon from
RRR9 (Behaviour) seems reduced by about a third of its initial value,
the ES heterogeneity for this phenomenon is rather small to begin with.
Using a significance test, we can not distinguish this value confidently
from zero. This implies that for the majority of phenomena, differences
in score reliability can not be made responsible for heterogeneity in
standardized effect sizes.

@@

*Note:* $\hat{\mu}_{\rho_{XX'}}$ is the back-transformed meta-analytic
estimate of mean score reliability; $\tau{\mu}_{\rho_{XX'}}$ is the
back-transformed estimate of heterogeneity in score reliability; Note
that, since in a meta-analysis, we attempt to estimate mean and variance
of the parameter's true distribution, in this table
$\hat{\tau}_{\delta_0}$ represents the estimate of heterogeneity in
uncorrected $d$; accordingly, $\hat{\tau}_{\delta}$ represents the
estimate of the heterogeneity in corrected $d_c$.

Secondly, in @tbl-tauES we find that for ten out of twelve phenomena, ES
heterogeneity grows as a result from the attenuation correction
procedure. Even though these differences tend to be small, it is clear
that in all those cases, not only did accounting for score reliability
not explain any ES heterogeneity, we find even more than initially
expected. This is incompatible with claims made in W&D and H&S, the
attenuation correction increased ES heterogeneity. Concerning the
remaining two phenomena, for the phenomenon described in ML5 - Shnabel
(RPP), the reduction in heterogeneity is miniscule and the estimate of
heterogeneity is not statistically significant in the first place. Most
likely, the difference in ES heterogeneity estimates can be explained by
issues of estimation, as only eight replications are available.
Similarly, only for the phenomenon described in RRR9 (Behaviour), we
find a seemingly substantial reduction in heterogeneity due to the
attenuation correction procedure. The estimates of ES heterogeneity are
not statistically significant, however, which hampers our ability to
meaningfully interpret and compare these estimates. For the majority of
phenomena it appears that differences in score reliability do not
inflate heterogeneity in uncorrected effect sizes. Instead these
differences mask ES heterogeneity, that can be unveiled by accounting
for score reliability.

## Relationship between score reliability and effect sizes in terms of heterogeneity

In the re-analysis of archival data in the previous section we
demonstrated that, at least for the multi-site direct replication and
PSA-projects, score reliability does not hold any noteworthy value as a
potential moderator explaining ES heterogeneity. Additionally, we
demonstrated that claims formulated in W&D and H&S regarding how score
reliability increases ES heterogeneity are false. We found a large
number of phenomena where, contrary to popular claims, correcting for
score reliability increases ES heterogeneity, instead of decreasing it.

@@

It might help to think about linear models, when discussing the
intuition that might underlie the idea, that controlling for score
reliability should reduce ES heterogeneity. Heterogeneity describes
unexplained variation in a parameter's true distribution, that we model
using meta-analytical techniques [@borenstein2009; @higgins2002]. In
that way, the idea of heterogeneity is closely related to residual
variance, as we observe it in a simple linear model. Both describe
variation that could not be explained by the predictive variables in the
case of the simple linear model, or not explained by incorporating the
Standard Errors in the case of a meta-analytic model.

For the simple linear model, if we add an additional predictor to the
model that we know shares variance with the parameter that we aim to
predict, this should reduce the amount of residual variance that we
observe across most instances. Generally, removing a source of variance
by controlling for it, we expect any unexplained variation to shrink.
Similarly, we know that differences in score reliability, by definition,
are related to differences in standardized effect sizes. Therefore,
intuitively, it seems sensible to expect that controlling for
differences in score reliability should reduce the unexplained variance
in standardized effect sizes, as described in H&S and W&D.

@@

We adjusted @eq-HS using the notation used throughout this text. Here,
$\delta_0$, carrying a 0 in the index, refers to the uncorrected ES,
undistorted by sampling error but not corrected for measuring error. On
the other hand $\delta$, no index, refers to the corrected ES,
undistorted by sampling error and measurement error. Lastly,
$\rho_{XX'}$ refers to score reliability. Hunter & Schmidt
[-@hunter2004] claim that @eq-HS demonstrates that differences in score
reliability inflate ES heterogeneity, which means that an attenuation
correction procedure would have to reduce ES heterogeneity. However, it
can be easily verified that @eq-HS does not allow for such claims.
Depending on both the mean value $\mu_{\rho_{xx‘}}$ and the
heterogeneity $\tau^2_{\rho_{xx‘}}$ in score reliability, ES
heterogeneity might be increased or decreased by the differences in
score reliability.

@@

Realising that an ES, as defined in @eq-d, is actually a ratio, allows
for a different analytical description of how differences in score
reliability can affect ES heterogeneity. If the phenomenon is
heterogeneous and score reliability varies across replications, it seems
sensible to assume that both numerator and denominator (MD and
$\sigma_X$) from @eq-d vary across replications. In that case, the ES
can be described as a random variable stemming from a ratio
distribution. A ratio distribution is a probability distribution
constructed by dividing one random variable by a second random variable:
$Z = \frac{X}{Y}$. First order taylor approximation may be used to
generate estimates of mean and variance of the ratio variable
[@vanKempen1999]. Taylor approximations typically require no specific
assumptions regarding the distribution form, but that the variable (here
a ratio) is a) differentiable, b) has existing moments and that c) its
higher-order moments are negligible (e.g. limited variability and
monotonicity) (still need a good reference here). Here we use the
Taylor-estimator of the ratio’s variance to demonstrate how differences
in score reliability may affect heterogeneity in ES.

@@

In @eq-rel we see that total standard deviation $\sigma_X$ is
constructed only from (the square of the sum of) error score variance
$\sigma_E^2$ and true score variance $\sigma_T^2$. Measurement
precision, since this should not pertain the true scores, should only
affect the error score variance. Therefore, if the heterogeneity in
$\sigma_X$ was introduced purely by differences in measurement
precision, an attenuation correction procedure can remove that
heterogeneity. Alternatively, the underlying latent variable may not be
distributed identically across replications. In that case, true score
variance $\sigma^2_T$ would vary across replications and induce
heterogeneity in $\sigma_X$. This heterogeneity would lead to
differences in standardized differences ES as well, but it can not be
removed by means of an attenuation correction procedure.

@@

The differences between the two [Equations @eq-ratio_d0] and
[-@eq-ratio_d] demonstrates how ES heterogeneity is affected by an
attenuation correction procedure. According to the quotes in H&S and
W&D, @eq-ratio_d0 should lead to a larger value of
$\tau^2\left[\delta_0\right]$, compared to the estimate of
$\tau^2\left[\delta\right]$ from @eq-ratio_d. However, as demonstrated
in the re-analysis of archival data, we can already expect that this can
not be an implication of these \[Equations @eq-ratio_d0\] and
[-@eq-ratio_d].

Instead, what is guaranteed is that the expected value of $\sigma_T$ is
smaller than the expected value of $\sigma_X$. Similarly, the
heterogeneity in $\sigma_T$ is guaranteed to be smaller than the
uncorrected heterogeneity in $\sigma_X$. However, since the expected
value of either standard deviation is placed in the denominator of the
function and the variance of either is placed in the numerator of the
function, @eq-HS is not sufficient to explain how heterogeneity in ES
changes due to the attenuation correction.

@@

As long as the inequality defined in @eq-ineq_init holds true, an
attenuation correction procedure would lead to larger estimates of ES
heterogeneity. Conditions under which @eq-ineq_init does not hold are
conditions under which the claims made in W&D and H&S are met.
Therefore, this inequality is essentially the opposite of the claims
made in W&D and H&S. Unfortunately, the inequality in this form does not
convey sufficient information about the conditions under which it can
potentially hold. In Appendix A, we rearranged @eq-ineq_init, so we end
up with a form that we can use to actually derive conditions under which
ES heterogeneity can be inflated or deflated by differences in score
reliability.

@@

$R_1$ describes the relative size of mean true score variance, compared
to the mean observed score variance. This informs us, how much of the
mean total score variance can be attributed to mean true score variance.
Therefore, this metric is close to the average score reliability. Even
though formally $R_1$ and the meta-analytic mean score reliability
$\mu_{\rho_{xx'}}$ do not share not the same value, in practice we
observe little to no differences between the two parameters. Large
values indicate that correcting the individual score variances (or
standard deviations for that matter) using the attenuation correction
procedure should lead to very little change in ES, as most of the
variation observed were due to genuine differences in true scores. Small
values indicate the opposite, the observed score variance was due to a
larger amount of random error score variance, leading to a stronger
correction.

@@

The metric $R_2$ on the other hand describes in how far the attenuation
correction procedure has successfully reduced heterogeneity in score
variance $\tau^2_{\sigma^2_X}$. A value of 1 indicates that the
heterogeneity in true score variance $\sigma^2_T$ is essentially just as
large as the heterogeneity initially observed. Smaller values indicate
how much heterogeneity is "left", after applying an attenuation
correction, relative to the initial heterogeneity. This means that a
value of .7 indicates that about 70% of heterogeneity in score variance
remains, even after correcting for differences in score reliability. As
only 30% of score variance heterogeneity could be removed, this would
imply that differences in error score variance $\sigma^2_E$ were
responsible for less than a third of the heterogeneity in score
variances found. On the contrary, more than two thirds of heterogeneity
could be attributed to actual differences in how the underlying true
scores are distributed across samples.

Concerning the re-arranged inequality from Appendix A, we identify one
necessary but not sufficient condition or rule described in
@eq-ineq_short.

@@

With necessary but not sufficient we mean that, as long as
@eq-ineq_short holds, an attenuation correction will inevitably increase
ES heterogeneity. However, if the inequality does not hold, it is not
guaranteed that the attenuation correction can decrease ES
heterogeneity. Whether ES heterogeneity can be decreased depends on
other parameters, that we left out of this short form, but can be found,
accompanied by a more detailed description, in @eq-ineq_fin.

@@

$R_1$ essentially describes the relative size of mean score variance
after the attenuation correction, and $R_2$ essentially describes the
relative size of score variance heterogeneity after the attenuation
correction. Therefore, @eq-ineq_short states the following regarding the
change in mean value and heterogeneity due to the correction: ES
heterogeneity will always be reduced, as long as the remaining
heterogeneity in score variance is not substantially smaller than the
size of the remaining mean value.

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$ largely
depends on the relative measurement precision. If measurements were made
more precisely, average score reliability and $R_1$ would be higher. On
the other hand, $R_2$ depends on how different the measurement precision
was at the individual administrations, but also how different the
individual administrations sites are regarding their true score
variance. If the administration sites carry large differences in true
score variance $R_2$ is unlikely to be substantially smaller than $R_1$.

@@

@tbl-tauES showed that for all phenomena, for which we identified
statistically significant ES heterogeneity, the heterogeneity was even
larger after the attenuation correction procedure. Here, we will
re-analyse the data, estimating $R_1$ and $R_2$ for each phenomenon
respectively. Therein, we not only show that our claims derived from
analytical arguments hold up in empirical assessment, but also
demonstrate how the metrics $R_1$ and $R_2$ may be used to inform how
score reliability affected ES heterogeneity.

@@

For the following analysis, Boot-Err is used. As Boot-Err makes use of
meta-analytic techniques and bootstrap sampling, we perform analyses
using the packages *metafor*, version 4.8-0 [@metafor], and *boot*,
version 1.3-31 [@boot], in the statistical programming language *R*,
version 4.4.2 [@RCore].

@@

@tbl-R1R2 reports the meta-analytic estimates of observed and true score
variances, including their heterogeneity and metrics $R_1$ and $R_2$
across all 12 phenomena. In @tbl-R1R2 we see that for ten out of 12
phenomena, the $R_2$ metric is larger than the $R_1$ metric.
Additionally, we observe that across eight phenomena, $R_2$ is larger
than .96. This implies that of the heterogeneity observed in score
variances, 97% or more could not be explained by differences in random
error introduced by the measuring instrument, but by genuine differences
in the underlying distribution.

For the two exceptions, those phenomena where the attenuation correction
procedure actually led to a decrease in ES heterogeneity (RRR9
(Behaviour) and ML5 - Shnabel (RPP)), the significance test for the ES
heterogeneity was not significant, both before and after the correction.
The results regarding $R_2$ indicate that the majority of score variance
heterogeneity was due to differences in true score variance
$\sigma^2_T$. Since either the heterogeneity was so small or the sample
size so low that the heterogeneity could not be statistically
distinguished from zero, we should be cautious in the interpretation of
a reduction in that estimate. Most likely, the reduction in
heterogeneity is better explained by imprecise estimation than an actual
reduction of heterogeneity.

Lastly, for two out of the 12 phenomena, the one described in ML5 -
Albarracin and the other described in ML5 - Shnabel (Rev), we find
$\tau_{\sigma^2_X}$ of 0. Therefore, the $R_2$ metric could not be
computed, as the denominator would have been zero. However, we also did
not identify statistically significant ES heterogeneity in the first
place. The phenomena were assessed with a particularly low number of
replications (9 and 8 respectively). Estimation precision for
heterogeneity in variance components, even more so than for ES, largely
depends on the sample size. Most likely, the low sample size led to
implausible estimates in both true and total score variance
heterogeneity and subsequently to an implausible estimate of $R_2$.
Meta-analytic estimates, especially those concerning score variance
components such as $R_2$, therefore should be interpreted with caution,
if the sample size was particularly low.

Generally, @tbl-R1R2 implies that for all measurements concerning these
phenomena, the heterogeneity in observed score variances could not be
substantially explained by differences in measurement quality. For all
phenomena, $R_2$ was larger than $R_1$, implying that an attenuation
correction procedure could not reduce ES heterogeneity in any of these
cases. The results in @tbl-tauES agree with this conclusion, as no
phenomena could be identified where statistically significant ES
heterogeneity was accompanied by a reduction in differences due to the
correction.

@@

We have examined meta-analytic heterogeneity in standardized mean
differences. Since score reliability is often discussed as a source for
such heterogeneity, we estimated the heterogeneity before and after
correcting for differences in reliability coefficients using an
attenuation correction procedure. We found that a) in the majority of
cases, the change in ES heterogeneity is small to non-existent. This
implies that, across the data-sets we got to examine, differences in
score reliability do not serve as an appropriate explanation for ES
heterogeneity. We also found that b) in the majority of cases, the
direction of change in ES heterogeneity does not align with assumptions
formulated in W&D and H&S. In almost all cases (every single case, if we
restrict ourselves to phenomena where we found statistically significant
differences in ES), the attenuation correction led to increased
estimates of ES heterogeneity. This means that the claims found in W&D
and H&S, implying that an attenuation correction can reduce ES
heterogeneity, stem from a general misunderstanding of how score
reliability can affect ESs.

@@

Since $R_1$ is carrying the exponent of 3, it implies that the reduction
in mean score variance is affecting the ES heterogeneity more strongly
than the reduction in score variance heterogeneity. This is best
described through an example: We know what an acceptable average score
reliability looks like in the area of Psychology. Researchers largely
target a score reliability of at least .7, preferably .8. An average
score reliability of .7 implies that 70% of the variance observed in
scores can actually be attributed to true underlying differences. From
our derived rule, we know that, as long as $R_2$ is larger or equal to
$.7^3 = .343$, the attenuation correction procedure can only increase ES
heterogeneity. If we want to explain ES heterogeneity through score
reliability, even though only 30% of the variance observed in score can
be attributed to random measurement error, at least 65.7%, almost two
thirds, of the heterogeneity in score variance would need to be
attributable to differences in the error score variance. In other words,
at most 34.3% of heterogeneity in score variances may be due to actual
differences in how the latent variable is distributed (in terms of
spread) across the administration sites.

In the examples we re-analysed throughout this text, as well as the
majority of research studies done in the behavioural sciences
[@hultsch2002; @bornstein2013], no proper sampling techniques are
employed. Most samples where these instruments are applied come to be
through convenience sampling, where students self-select into studies.
Without a valid sampling scheme, we essentially have have no basis to
assume anything about the true score variance [@hultsch2002]. Most
importantly, however, we have no reason to assume that the true score
variance could potentially be homogeneous or stable across
administration sites. Instead, it is very likely that the convenience
sampling procedures produce samples that vary extremely strongly in how
they are put together, leading to large degrees of true score variance
heterogeneity. At the same time, score reliability is a low-hanging
fruit to optimise as most researchers are aware and trained in assessing
score reliability [@hussey2025].

>>>>

The estimates concerning the meta-analytic mean and heterogeneity of
these standardized effect sizes attempt to describe their population
distribution. Therefore, over the following paragraphs, whenever we
refer to estimates derived by means of random effects meta-analysis, we
will use the Greek letter $\delta_0$ to denote estimates of uncorrected
$d$'s population distribution and $\delta$ to denote estimates of
corrected $d_c$'s population distribution.

@@

In @eq-rel_back_mu, $\mu_{\rho_{XX'}}$ is the back-transformed
meta-analytic mean score reliability - on the original coefficient's
scale between 0 and 1. Similarly, $\tau_{\mu_{\rho_{XX'}}}$ is the
back-transformed heterogeneity in score reliability:

@@

@tbl-tauES contains the results from performing an RG-MA on the
Cronbach's Alpha coefficients of the scales used to measure the 12
remaining phenomena. Additionally, it contains estimates of
heterogeneity for both uncorrected ES ($\delta_0$) and corrected ES
($\delta$), which are central to assess the research question posed in
section *Attenuation Correction*. Most importantly, @tbl-tauES
demonstrates that for all but one of the phenomena, the attenuation
correction procedure barely has an impact on the extent of ES
heterogeneity. However, while the heterogeneity for the phenomenon from
RRR9 (Behaviour) seems to be reduced by about a third of its initial
value, the ES heterogeneity for this phenomenon is rather small to begin
with. Using a significance test, we can not distinguish this value
confidently from zero. This implies that for the majority of phenomena,
differences in score reliability can not be made responsible for
heterogeneity in standardized effect sizes.

@@

*Note:* $\hat{\mu}_{\rho_{XX'}}$ is the back-transformed meta-analytic
estimate of mean score reliability; $\tau_{\mu_{\rho_{XX'}}}$ is the
back-transformed estimate of heterogeneity in score reliability;
$\hat{\tau}_{\delta_0}$ represents the estimate of heterogeneity in
uncorrected $d$; $\hat{\tau}_{\delta}$ represents the estimate of the
heterogeneity in corrected $d_c$.

Secondly, in @tbl-tauES we find that for ten out of twelve phenomena, ES
heterogeneity grows as a result from the attenuation correction
procedure. However, it needs to be noted that these differences are small
and could not be distinguished from sampling uncertainty in a statistical
test. For this reason, we do not mean to interpret these differences, while 
we note that the direction of differences are clearly inconsistent with
expectations formulated in the literature discussed so far.
These results are incompatible with claims made in W&D and H&S.

Besides the direction of these differences, it is noteworthy that considering
differences in score reliability barely affects heterogeneity in standardized
effect sizes. [ausführen, oder kommt das noch einmal wieder?]

Concerning the
remaining two phenomena, for the phenomenon described in ML5 - Shnabel
(RPP), the reduction in heterogeneity is miniscule and the estimate of
heterogeneity is not statistically significant in the first place. Most
likely, the difference in ES heterogeneity estimates can be explained by
issues of estimation, as only eight replications are available.
Similarly, only for the phenomenon described in RRR9 (Behaviour), we
find a seemingly substantial reduction in heterogeneity due to the
attenuation correction procedure. The estimates of ES heterogeneity are
not statistically significant, however, which hampers our ability to
meaningfully interpret and compare these estimates. For the majority of
phenomena it appears that differences in score reliability do not
inflate heterogeneity in uncorrected effect sizes. Instead these
differences mask ES heterogeneity, that can be unveiled by accounting
for score reliability.

## Relationship between score reliability and effect size heterogeneity

In the re-analysis of archival data in the previous section we
demonstrated that, at least for the multi-site direct replication and
PSA-projects, score reliability does not hold any noteworthy value as a
potential moderator explaining ES heterogeneity. Additionally, we
demonstrated that claims formulated in W&D and H&S regarding how score
reliability heterogeneity increases ES heterogeneity are false. We found a large
number of phenomena where, contrary to popular claims, correcting for
score reliability increases ES heterogeneity, instead of decreasing it.

@@

It might help to think about linear models, when discussing the
intuitive idea that controlling for score
reliability should reduce ES heterogeneity. Heterogeneity describes
unexplained variation in a parameter's true distribution, that we model
using meta-analytical techniques [@borenstein2009; @higgins2002]. In
that way, the idea of heterogeneity is closely related to residual
variance, as we observe it in a simple linear model. Both describe
variation that could not be explained by the model specifications, be it a model that 
employs predictive variables like a simple linear model or a meta-analytic
model incorporating uncertainty in terms of standard errors.

The residual variance that remains in a simple linear model will be reduced
in all instances, where we use a predictor that we know shares 
variance with the predicted parameter. Generally, removing a source of variance
by controlling for it, we expect any unexplained variation to shrink.
Similarly, we know that differences in score reliability, by definition,
are related to differences in standardized effect sizes. Therefore,
intuitively, it seems sensible to expect that controlling for
differences in score reliability should reduce the unexplained variance
in standardized effect sizes, as described in H&S and W&D.

@@

We adjusted @eq-HS using the notation used throughout this text. Here,
$\delta_0$, carrying a 0 in the index, refers to the uncorrected ES,
undistorted by sampling error but not corrected for measuring error. Without an 
index, $\delta$, on the other hand, refers to the corrected ES,
undistorted by sampling error and measurement error. Lastly,
$\rho_{XX'}$ refers to score reliability. Hunter & Schmidt
[-@hunter2004] claim that @eq-HS demonstrates that differences in score
reliability inflate ES heterogeneity, which means that an attenuation
correction procedure would have to reduce ES heterogeneity. However, it
can be easily verified that @eq-HS does not allow for such claims.
Depending on both the mean value $\mu_{\rho_{xx‘}}$ and the
heterogeneity $\tau^2_{\rho_{xx‘}}$ in score reliability, ES
heterogeneity might be increased or decreased by the differences in
score reliability.

@@

Realising that an ES, as defined in @eq-d, is actually a ratio, allows
for a different analytical description of how differences in score
reliability can affect ES heterogeneity. If the phenomenon is
heterogeneous and score reliability varies across replications, it seems
sensible to assume that both numerator and denominator (MD and
$\sigma_X$) from @eq-d vary across replications. In that case, the ES
can be described as a random variable stemming from a ratio
distribution. A ratio distribution is a probability distribution
constructed by dividing one random variable by a second random variable:
$Z = \frac{X}{Y}$. First order taylor approximation may be used to
generate estimates of mean and variance of the ratio variable
[@vanKempen1999]. Taylor approximations typically require no specific
assumptions regarding the distribution form, but that the variable (here
a ratio) is a) differentiable, b) has existing moments and that c) its
higher-order moments are negligible (e.g. limited variability and
monotonicity) (Asymptotic statistics, A.W. van der Vaart - find book online). Here we use the
Taylor-estimator of the ratio’s variance to demonstrate how differences
in score reliability may affect heterogeneity in ES.

@@

In @eq-rel we see that total standard deviation $\sigma_X$ is
constructed only from error score variance
$\sigma_E^2$ and true score variance $\sigma_T^2$. Measurement
precision should generally only
affect the error score variance. Therefore, if the heterogeneity in
$\sigma_X$ was introduced purely by differences in measurement
precision, an attenuation correction procedure can remove that
heterogeneity. Alternatively, the underlying latent variable may not be
distributed identically across replications. In that case, true score
variance $\sigma^2_T$ would vary across replications and induce
heterogeneity in $\sigma_X$. This heterogeneity would lead to
differences in standardized differences ES as well, but it can not be
removed by means of an attenuation correction procedure.

@@

The differences between the two [Equations @eq-ratio_d0] and
[-@eq-ratio_d] demonstrates how ES heterogeneity is affected by an
attenuation correction procedure. According to the quotes in H&S and
W&D, @eq-ratio_d0 should lead to a larger value of
$\tau^2\left[\delta_0\right]$, compared to the estimate of
$\tau^2\left[\delta\right]$ from @eq-ratio_d. However, as demonstrated
in the re-analysis of archival data, we can already expect that this can
not be an implication of these Equations.

Instead, what is guaranteed is that the expected value of $\sigma_T$ is
smaller than or equal to the expected value of $\sigma_X$. Similarly, the
heterogeneity in $\sigma_T$ is guaranteed to be smaller than or equal to the
uncorrected heterogeneity in $\sigma_X$. However, since the expected
value of either standard deviation is placed in the denominator of the
function and the variance of either is placed in the numerator of the
function, @eq-HS is not sufficient to explain how heterogeneity in ES
changes due to the attenuation correction.

@@

As long as the inequality defined in @eq-ineq_init holds true, an
attenuation correction procedure would lead to larger estimates of ES
heterogeneity. Conditions under which @eq-ineq_init does not hold are
conditions under which the claims made in W&D and H&S are met.
Unfortunately, the inequality in this form does not
convey sufficient information about the conditions under which it can
potentially hold. In Appendix A, we rearranged @eq-ineq_init, so we end
up with a form that we can use to actually derive conditions under which
ES heterogeneity can be inflated or deflated by differences in score
reliability.

@@

$R_1$ describes the relative size of mean true score variance, compared
to the mean observed score variance. This informs us, how much of the
mean total score variance can be attributed to mean true score variance.
Therefore, this metric is close to the average score reliability. Even
though formally $R_1$ and the meta-analytic mean score reliability
$\mu_{\rho_{xx'}}$ do not share not the same value, in practice we
observe little to no differences between the two parameters. [add to appendix]. Large
values indicate that correcting the individual score variances (or
standard deviations for that matter) using the attenuation correction
procedure should lead to very little change in ES, as most of the
variation observed were due to genuine differences in true scores. Small
values indicate the opposite, the observed score variance was due to a
larger amount of random error score variance, leading to a stronger
correction. The metric $R_2$ on the other hand describes in how far the attenuation
correction procedure has successfully reduced heterogeneity in score
variance $\tau^2_{\sigma^2_X}$.

@@

A value of 1 indicates that the
heterogeneity in true score variance $\sigma^2_T$ is exactly as
large as the heterogeneity initially observed. Smaller values indicate
how much heterogeneity is "left", after applying an attenuation
correction, relative to the initial heterogeneity. This means that a
value of .7 indicates that about 70% of heterogeneity in score variance
remains, even after correcting for differences in score reliability. As
only 30% of score variance heterogeneity could be removed, this would
imply that differences in error score variance $\sigma^2_E$ were
responsible for less than a third of the heterogeneity in score
variances found. On the contrary, more than two thirds of heterogeneity
could be attributed to actual differences in how the underlying true
scores are distributed across samples.

Concerning the re-arranged inequality from Appendix A, we identify one
sufficient but not necessary condition or rule described in
@eq-ineq_short.

@@

With sufficient but not necessary we mean that, as long as
@eq-ineq_short holds, an attenuation correction will inevitably increase
ES heterogeneity. However, if the inequality does not hold, it is not
guaranteed that the attenuation correction can decrease ES
heterogeneity. Whether ES heterogeneity can be decreased depends on
additional parameters, that we left out of this short form, but can be found,
accompanied by a more detailed description, in @eq-ineq_fin, Appendix A.

@@

$R_1$ essentially describes the relative size of mean score variance
after the attenuation correction, and $R_2$ essentially describes the
relative size of score variance heterogeneity after the attenuation
correction. Therefore, @eq-ineq_short states the following regarding the
change in mean value and heterogeneity due to the correction: ES
heterogeneity will always be increased, as long as the remaining proportion of
heterogeneity in score variance is not substantially smaller than the
remaining proportion of its mean value.

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$ largely
depends on the relative measurement precision. If measurements were made
more precisely, average score reliability and $R_1$ would be higher. On
the other hand, $R_2$ depends on how different the measurement precision
was at the individual administrations, but also how different the
individual administrations sites are regarding their true score
variance. If the administration sites carry large differences in true
score variance $R_2$ is unlikely to be substantially smaller than $R_1$. [Na ja, eine Reliabilität von .9 bedeutet erstmal, dass es 9mal so viel wahre Varianz wie Fehlervarianz gibt. In diesem Fall müsste R2 unter 0.73 liegen. Das impliziert in der Heterogenität ein Verhältnis von ungefähr 3 : 1. Die 9mal größere wahre Varianz müsste also für weniger als ¾ der Heterogenität verantwortlich sein. In Deinem Textbeispiel (Reliabilität von .8) müsste die 4mal größere wahre Varianz für weniger als die Hälfte der Heterogenität verantwortlich sein. – Das erscheint mir beides per se ziemlich unplausibel… Also auch dann, wenn die „administration sites do NOT carry large differences in true score variance”. – In anderen Worten: Die Frage ist, ob die absolute Größe der Varianz überhaupt irgendeine Rolle spielt oder ob alles lediglich an den Verhältnissen hängt, die durch die Reliabilität vorgegeben sind. Ich habe dazu aber auch keine eindeutige Antwort… Wenn ich überzeugt wäre, dass sich *homogene* true score variances durch planvolle, kontrollierte Stichprobenziehung herstellen lassen, wäre das ein Gegenargument.]

@@

@tbl-tauES showed that for all phenomena, for which we identified
statistically significant ES heterogeneity, the heterogeneity was even
larger after the attenuation correction procedure. Here, we will
re-analyse the data, estimating $R_1$ and $R_2$ for each phenomenon
respectively. Therein, we not only show that our claims derived from
analytical arguments hold up in empirical assessment, but also
demonstrate how the metrics $R_1$ and $R_2$ help understand how
score reliability affected ES heterogeneity in practice.

@@

As Boot-Err makes use of
meta-analytic techniques and bootstrap sampling, we perform analyses
using the packages *metafor*, version 4.8-0 [@metafor], and *boot*,
version 1.3-31 [@boot], in the statistical programming language *R*,
version 4.4.2 [@RCore].

@@

@tbl-R1R2 reports the meta-analytic estimates of observed and true score
variances, including their heterogeneity and metrics $R_1$ and $R_2$
across all 12 phenomena. We see that for ten out of 12
phenomena, the $R_2$ metric is larger than the $R_1$ metric.
Additionally, we observe that across eight phenomena, $R_2$ is larger
than .96. This implies that of the heterogeneity observed in score
variances, 97% or more could not be explained by differences in random
error introduced by the measuring instrument, but by genuine differences
in the underlying distribution.

For the two exceptions in @tbl-tauES, those phenomena where the attenuation correction
procedure actually led to a decrease in ES heterogeneity (RRR9
(Behaviour) and ML5 - Shnabel (RPP)), we find that $R_2$ is larger than $R_1$.
This clearly contradicts our interpretation of @eq-ineq_int. As long as $R_2$ is
larger than $R_1^3$, we expect a correction procedure to increase effect 
size heterogeneity. However, we recommend to be cautious when interpreting both the change and
the direction of said change for these two projects. 
The significance test for the ES
heterogeneity was not significant, both before and after the correction.
Since either the heterogeneity was so small or the sample
size so low that the heterogeneity could not be statistically
distinguished from zero, most likely, the reduction in
heterogeneity is better explained by imprecise estimation than an actual
reduction of heterogeneity.

Lastly, for two out of the 12 phenomena, ML5 -
Albarracin and ML5 - Shnabel (Rev), we find
$\tau_{\sigma^2_X}$ of 0. Therefore, the $R_2$ metric could not be
computed, as the denominator would have been zero. However, we also did
not identify statistically significant heterogeneity in score reliability
across replications, nor ES heterogeneity. The phenomena were assessed with a 
particularly low number of
replications (9 and 8 respectively). Estimation precision for
heterogeneity in variance components, even more so than for ES, largely
depends on the sample size. Two alternative explanations for this observation
are available: either, there is no heterogeneity in score variances across 
replications. In that case, there is also no heterogeneity that could be
broken down into true and error score variance heterogeneity. Alternatively,
the low sample size led to
implausible estimates in both true and total score variance
heterogeneity and subsequently to an implausible estimate of $R_2$.
Meta-analytic estimates, especially those concerning score variance
components such as $R_2$, should generally be interpreted with caution,
if the sample size was particularly low.

Generally, @tbl-R1R2 implies that for none of the measurements concerning these
phenomena, the heterogeneity in observed score variances could be
substantially explained by differences in measurement quality. For all
phenomena, $R_2$ was larger than $R_1$, implying that an attenuation
correction procedure could not reduce ES heterogeneity in any of these
cases. The results in @tbl-tauES agree with this conclusion, as no
phenomena could be identified where statistically significant ES
heterogeneity was accompanied by a reduction in differences due to the
correction.

@@

We have examined meta-analytic heterogeneity in standardized mean
differences. Since score reliability is often discussed as a source for
such heterogeneity, we estimated the heterogeneity before and after
correcting for differences in reliability coefficients using an
attenuation correction procedure. While we initially observed heterogeneity
in score reliability across almost all measuring scales, the extent of this
heterogeneity was very small across all scales. This initial finding already
implies that differences in score reliability can only have a tiny impact
on ES heterogeneity, if any.

This implication fits our subsequent finding, as we found that a) in the majority of
cases, the change in ES heterogeneity is small to non-existent. Across the 
data-sets we got to examine, differences in
score reliability do not serve as an appropriate explanation for ES
heterogeneity. We also found that b) in the majority of cases, the
direction of change in ES heterogeneity does not align with assumptions
formulated in W&D and H&S. In almost all cases (every single case, if we
restrict ourselves to phenomena where we found statistically significant
differences in ES), the attenuation correction led to increased
estimates of ES heterogeneity. This means that the claims found in W&D
and H&S, implying that an attenuation correction will reduce ES
heterogeneity, stem from a general misunderstanding of how score
reliability can affect ESs.

@@

Since $R_1$ is carrying the exponent of 3, it implies that the reduction
in mean score variance is affecting the ES heterogeneity more strongly
than the reduction in score variance heterogeneity. This is best
described through an example: We know what an acceptable average score
reliability looks like in the area of Psychology. Researchers largely
target a score reliability of at least .7, preferably .8. An average
score reliability of .7 implies that 70% of the variance observed in
scores can actually be attributed to true underlying differences. From
our derived rule, we know that, as long as $R_2$ is larger or equal to
$.7^3 = .343$, the attenuation correction procedure can only increase ES
heterogeneity. If we want to explain ES heterogeneity through score
reliability, even though only 30% of the variance observed in score can
be attributed to random measurement error, at least 65.7%, almost two
thirds, of the heterogeneity in score variance would need to be
attributable to differences in the error score variance. In other words,
at most 34.3% of heterogeneity in score variances may be due to actual
differences in how the latent variable is distributed (in terms of
spread) across the administration sites. This disparity grows even larger, 
the higher the average score reliability is.

In the 
majority of research studies done in the social and behavioural sciences
[@hultsch2002; @bornstein2013; @ashraf2017; @sears1986; @zhao2021], no 
proper sampling techniques beyond convenience sampling are
employed. Most samples re-analysed throughout this text came to be
through convenience sampling, where students self-select into studies.
Without a valid sampling scheme, we essentially have no basis to
assume anything about the true score variance [@hultsch2002]. Most
importantly, however, we have no reason to assume that the true score
variance could potentially be homogeneous or stable across
administration sites. Instead, it is very likely that the convenience
sampling procedures produce samples that vary extremely strongly in how
they are put together, leading to large degrees of true score variance
heterogeneity. At the same time, score reliability is a low-hanging
fruit to optimise as most researchers are aware and trained in assessing
score reliability [@hussey2025].
